Design and implemention of open health data systems with standardized data and analytics capabilities for low- and middle income countries

### 2024-08-29 index.qmd
Editorial suggestions

To achieve scaleability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support improvements in the wider health sector along five dimensions, namely [@kelley2020digital]:

@@

Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions are within reach. As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs. By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].

MomCare is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS. For example, to implement the value-based healthcare business logic of MomCare, detailed analysis of patient journeys is required. This functionality is currently not available as a standard; standardized reports that are available, such as DHIS2, contained insufficient information and analytical functions to support intervention that aim to improve continuity of care and adherence to clinical guidelines. As such, a large part of the day-to-day operation of MomCare revolved performaing data and analytics: data acquisition, data integration, analysis etc.

This situation is not unique to MomCare. In fact, many in the global digital health community think that data & analytics services should be an essential capibility of OHDSs going forward. This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper]. In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond it original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies etc. If we are to use the OpenHIE framework for secondary use, supported by data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.

@@

The main contribution of this paper is to propose how recent standards and open source implementations from the data engineering community can be integrated into the OpenHIE framework. In the following, we first describe how the lakehouse design pattern, being the most widely used data & analytics solution architecture, can be integrated in OpenHIE. To demonstrate the feasibility of this design, we present a proof-of-concept impelementation using open source technologies within the context of the MomCare programme. Code and digital artifacts of this demonstrator is available as supplementary material [TO DO: include links to support GitHub repositories]. Subquently, we compare this solution design with two widely used and operational OpenHIE-compliant open source frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/). Finally, we discuss our findings and propose routes for future development.

@@

The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack. However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.  Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form. The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.

@@

Clearly, a more open and scaleable platform was required if MomCare was to be implemented in more regions. This need led to a redesign of the underlying technical infrastructure of the MomCare project. The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing. First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard. Using the longitudinal dataset from approximately 28 thousand patient records, FHIR transformations script were developed and deployed using the mediator function of the IOL. The data was transformed into 10 FHIR v4 resources and the conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir]. The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations. For patients, either the mobile phone number or the healthcare insurance number was taken, depending on availability. A combination of name, address and latitude/longitude coordinates were used to uniquely identify organizations and locations, as Tanzania does not have a system in place for this purpose.

The second objective was to reproduce existing analytic reports, using the bulk FHIR data format as input. Here, the focus was to standardize the logic required for producing metrics and reports. The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function. Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library. Cross checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results. For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.

@@

@fig-openhim-platform shows a schematic overview of two data stacks that are supported in the OpenHIM platform. The Shared Health Record (SHR, implemented with HAPI FHIR server) and the Client Registry (CR, implemented with JeMPI server) are the sources that store clinical FHIR data and patient demographic data, respectively. The default data stack is based on streaming ingestion using Kafka into a Clickhouse database. As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics using a generic Kafka utility component. Subsequently, each FHIR resource topic is flatted with Kafka mappers that use FHIRPath. Superset is used as the tool for consuming the data to create dashboard visualizations.

@@

- Pattern of flattening FHIR resources with FHIRPath expressions is very close to the idea of SQL-on-FHIR. Although it doesn't adhere to this new standard in the strict sense, the philosophy of generating tabular views is the same
- When using the ELK stack, flattening is done at the end. Implementations of FHIRPath support Elasticsearch as an execution engine, also here 
- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scaleability is limited.

>>>>

To achieve scalability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support improvements in the wider health sector along five dimensions, namely [@kelley2020digital]:

@@

Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions are within reach. As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs. By actively encouraging pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].

MomCare is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS. For example, to implement the value-based healthcare business logic of MomCare, detailed analysis of patient journeys is required. This functionality is currently not available in a standard form; standardized reports that are available, such as DHIS2, contained insufficient information and analytical functions to support intervention that aim to improve continuity of care and adherence to clinical guidelines. As such, a large part of the day-to-day operation of MomCare involved performing data and analytics tasks: data acquisition, data integration, analysis etc.

This situation is not unique to MomCare. In fact, many in the global digital health community think that data & analytics services should be an essential capability of OHDSs going forward. This is exemplified by the adoption of the OpenHIE reference architecture [@openhie] by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper]. In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond its original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies, etc. If we are to use the OpenHIE framework for secondary use, supported by data and analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so. The lack of detailed specifications and consensus for this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.

@@

The main contribution of this paper is to propose how recent standards and open source implementations from the data engineering community can be integrated into the OpenHIE framework. In the following, we first describe how the lakehouse design pattern, being the most widely used data & analytics solution architecture, can be integrated in OpenHIE. To demonstrate the feasibility of this design, we present a proof-of-concept implementation using open source technologies within the context of the MomCare programme. Code and digital artifacts of this demonstrator is available as supplementary material [TO DO: include links to support GitHub repositories]. Subsequently, we compare this solution design with two widely used and operational OpenHIE-compliant open source frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/). Finally, we discuss our findings and propose routes for future development.

@@

The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack. However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.  Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form. The task of authoring these transformations and projections is not trivial and there are currently no standard mechanisms to support reuse.

@@

Clearly, a more open and scalable platform was required if MomCare was to be implemented in more regions. This need led to a redesign of the underlying technical infrastructure of the MomCare project. The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing. First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard. Using the longitudinal dataset from approximately 28 thousand patient records, FHIR transformations script were developed and deployed using the mediator function of the IOL. The data was transformed into 10 FHIR v4 resources and the conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir]. The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations. For patients, either the mobile phone number or the healthcare insurance number was taken, depending on availability. A combination of name, address and latitude/longitude coordinates were used to uniquely identify organizations and locations, as Tanzania does not have a system in place for this purpose.

The second objective was to reproduce existing analytic reports, using the bulk FHIR data format as input. Here, the focus was to standardize the logic required for producing metrics and reports. The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function. Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library. Cross-checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results. For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.

@@

@fig-openhim-platform shows a schematic overview of two data stacks that are supported in the OpenHIM platform. The Shared Health Record (SHR, implemented with HAPI FHIR server) and the Client Registry (CR, implemented with JeMPI server) are the sources that store clinical FHIR data and patient demographic data, respectively. The default data stack is based on streaming ingestion using Kafka into a Clickhouse database. As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics using a generic Kafka utility component. Subsequently, each FHIR resource topic is flattened with Kafka mappers that use FHIRPath. Superset is used as the tool for consuming the data to create dashboard visualizations.

@@

- Pattern of flattening FHIR resources with FHIRPath expressions is very close to the idea of SQL-on-FHIR. Although it doesn't adhere to this new standard in the strict sense, the philosophy of generating tabular views is the same
- When using the ELK stack, flattening is done at the end. Implementations of FHIRPath support Elasticsearch as an execution engine, also here 
- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scalability is limited.
